---
title: "Photographic_Bleaching"
author: "EL Strand"
date: "8/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading in required libraries. 
```{r}
library("vegan")
library("tidyverse")
library("lme4")
library("plotrix")
library("EnvStats")
library("car")
library("MuMIn")
```

Reading in quality controlled color data.
###Protocol for ImageJ Bleaching Score Analysis found ###[here](https://emmastrand.github.io/EmmaStrand_Notebook/ImageJ-Analysis-Protocols/).  
```{r}
Data <- read.csv("Data/Bleaching_ImageJ_QC.csv", header=T, sep=",", na.string="NA") #read in data file
Data <-na.omit(Data) # removing all rows with NA values
Data$SP.Time.ID <- paste0(Data$Species, "_",Data$Timepoint, "_", Data$PLUG.ID)# generating a grouping factor
```

Normalize coral color to color standards. 
```{r}
Data$Red.Norm.Coral <- Data$Red.Coral/Data$Red.Standard #normalize to standard
Data$Green.Norm.Coral <- Data$Green.Coral/Data$Green.Standard #normalize to standard
Data$Blue.Norm.Coral <- Data$Blue.Coral/Data$Blue.Standard #normalize to standard
```

Principal components analysis to reduce color score to single value representing the majority of the variance
```{r}

score <- Data %>% select("SP.Time.ID", "PLUG.ID", "Species", "Temperature", "CO2", "Timepoint", "Red.Norm.Coral","Green.Norm.Coral", "Blue.Norm.Coral") #select columns from dataframe
blch.scor <- score  %>% select("Red.Norm.Coral","Green.Norm.Coral", "Blue.Norm.Coral") #subset dataframe for score matrix only
blch.scor <- as.matrix(blch.scor) #create matrix
rownames(blch.scor) <- score$SP.Time.ID #name columns in matrix

dist <- vegdist(blch.scor, method="euclidean") #calculate distance matrix of color scores

PCA.color <- princomp(dist) #run principal components analysis
PCA.color # view variance explained by PCs

Blch <- as.data.frame(-PCA.color$scores[,1]) #extract PC1
Blch$SP.Time.ID <- rownames(blch.scor) # add the sample names to the bleahcing score

Blch  <- left_join(Blch, Data, by="SP.Time.ID") #make a dataframe of PC1 and experiment factors
colnames(Blch) #view column names
names(Blch)[1] <- "Bleaching.Score" #name column

#set levels of time factor
Blch$Timepoint <- factor(Blch$Timepoint, levels = c("Week1", "Week2", "Week3", "Week4", "Week5", "Week6", "Week7", "Week8", "Week9", "Week10", "Week11", "Week12", "Week13", "Week14", "Week15", "Week16")) 

Blch$group <- paste0(Data$Temperature,"_", Data$CO2, "_",Data$Timepoint)

```

### Plot data for outliers and Write out Color Score dataframe 
```{r}

Blch.meanse <- Blch %>% group_by(Species, Treatment, Timepoint) %>% #group data by factors
  summarise(mean = mean(Bleaching.Score), #calculate mean by group
            sem = std.error(Bleaching.Score)) #calculate sem by group

# Assign labels to use for figure
species.names <- list(
  'Mcapitata'="Montipora capitata",
  'Pacuta'="Pocillopora acuta")
species_labeller <- function(variable,value){
  return(species.names[value])}

cols <- c("lightblue", "blue", "salmon", "red3") #set colors

Blch_plot1 <- ggplot(Blch.meanse, aes(x=Timepoint, y = mean, group = Treatment, color = Treatment)) +
  geom_line() + geom_point() + #plot points and lines
  facet_grid(~Species, scales = "free", labeller = species_labeller) + #facet by Species
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.2) + #plot sem error bars
  scale_color_manual(values = cols) + #color points
  xlab("Timepoint") + #label x axis
  ylab("Color Score") + # label y axis
  theme_classic() + #set theme
  theme(legend.position = c(0.1,0.3)) + #set legend position
  theme(legend.text = element_text(size = 7)) + #set legend text size
  theme(legend.title = element_text(size = 9)) + #set legend text title size
  geom_vline(xintercept = c(7.8), colour="black", linetype="dotted") + #add vertical line
  theme(plot.title = element_text(face = 'bold.italic', size = 12, hjust = 0)) +  #set font
  scale_x_discrete(breaks = c("Week1", "Week2", "Week4", "Week6", "Week8", "Week12", "Week16"), labels = c("1 week", "2 week", "4 week", "6 week", "8 week", "12 week", "16 week")) + #set breaks
  theme(axis.text.x = element_text(angle = 60, vjust = 1.2, hjust = 1.3)) + #Set the text angle
  theme(panel.border = element_blank(), #set theme
  panel.grid.major = element_blank(),  #set theme
  panel.grid.minor = element_blank(),  #set theme
  # Change axis line
  axis.line = element_line(colour = "black")) 
Blch_plot1 #view plot

#remove outliers
Blch <- Blch %>% filter(Timepoint != "Week5") #remove week5 outliers

#remove Pacuta after week 12 for small sample size
Blch <- Blch %>% filter(Species != "Pacuta" | Timepoint != "Week13")
Blch <- Blch %>% filter(Species != "Pacuta" | Timepoint != "Week14")
Blch <- Blch %>% filter(Species != "Pacuta" | Timepoint != "Week15")
Blch <- Blch %>% filter(Species != "Pacuta" | Timepoint != "Week16")

Blch.meanse <- Blch %>% group_by(Species, Treatment, Timepoint) %>% #group data by factors
  summarise(mean = mean(Bleaching.Score), #calculate mean by group
            sem = std.error(Bleaching.Score)) #calculate sem by group

Blch_plot <- ggplot(Blch.meanse, aes(x=Timepoint, y = mean, group = Treatment, color = Treatment)) +
  geom_line() + geom_point() + #plot points and lines
  facet_grid(~Species, scales = "free", labeller = species_labeller) + #facet by Species
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.2) +  #plot sem error bars
  scale_color_manual(values = cols) + #color points
  xlab("Timepoint") + #label x axis
  ylab("Color Score") + # label y axis
  theme_classic() + #set theme
  theme(legend.position = c(0.1,0.3)) + #set legend position
  theme(legend.text = element_text(size = 7)) + #set legend text size
  theme(legend.title = element_text(size = 9)) + #set legend text title size
  geom_vline(xintercept = c(7.8), colour="black", linetype="dotted") + #add vertical line
  theme(plot.title = element_text(face = 'bold.italic', size = 12, hjust = 0)) + 
  scale_x_discrete(breaks = c("Week1", "Week2", "Week4", "Week6", "Week8", "Week12", "Week16"), labels = c("1 week", "2 week", "4 week", "6 week", "8 week", "12 week", "16 week")) + #set breaks
  theme(axis.text.x = element_text(angle = 60, vjust = 1.2, hjust = 1.3)) + #Set the text angle
  theme(panel.border = element_blank(),#set theme
  panel.grid.major = element_blank(),#set theme
  panel.grid.minor = element_blank(),#set theme
  # Change axis line
  axis.line = element_line(colour = "black"))+ 
  theme(strip.text = element_text(face = "bold.italic")) 
Blch_plot #view plot


write.table(Blch, file = "Output/Bleaching_Score_Output.csv", append = FALSE, quote = TRUE, sep = ",", eol = "\n", na = "NA", dec = ".", row.names = TRUE, col.names = TRUE, qmethod = c("escape", "double"), fileEncoding = "")


```

Analyze Color Score using Mixed effect modeling 
```{r}
Blch$PLUG.ID <- as.character(Blch$PLUG.ID) 
Blch$Timepoint <- as.factor(Blch$Timepoint) #set factors
mcap.color <- Blch %>% filter(Species == "Mcapitata") #subset by species
pacuta.color <- Blch %>% filter(Species == "Pacuta") #subset by species

hist(pacuta.color$Bleaching.Score) #view data distribution
hist(mcap.color$Bleaching.Score) #view data distribution
# Data is heavily left skewed and calls for transformation.
```

Testing Transformations for M.capitata
```{r}
#Our random factor is `(1|PLUG.ID)` because fragments were repeatedly measured
Mcap_Color_sqrt <- lmer(((Bleaching.Score+500)^(1/2)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=mcap.color) 
par(mfrow = c(2, 2))
hist(residuals(Mcap_Color_sqrt)) #view distribution
boxplot(residuals(Mcap_Color_sqrt)) #view distribution
qqPlot(residuals(Mcap_Color_sqrt)) #view qq
plot(fitted(Mcap_Color_sqrt), residuals(Mcap_Color_sqrt)) #view variance

Mcap_Color_qrt <- lmer(((Bleaching.Score+500)^(1/4)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=mcap.color) 
par(mfrow = c(2, 2))
hist(residuals(Mcap_Color_qrt))#view distribution
boxplot(residuals(Mcap_Color_qrt))#view distribution
qqPlot(residuals(Mcap_Color_qrt))#view qq
plot(fitted(Mcap_Color_qrt), residuals(Mcap_Color_qrt))#view variance

Mcap_Color_log <- lmer(log10((Bleaching.Score+500)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=mcap.color)
par(mfrow = c(2, 2))
hist(residuals(Mcap_Color_log))#view distribution
boxplot(residuals(Mcap_Color_log))#view distribution
qqPlot(residuals(Mcap_Color_log))#view qq
plot(fitted(Mcap_Color_log), residuals(Mcap_Color_log))#view variance

Mcap_Color_sq <- lmer(((Bleaching.Score+500)^2) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=mcap.color)
par(mfrow = c(2, 2))
hist(residuals(Mcap_Color_sq))#view distribution
boxplot(residuals(Mcap_Color_sq))#view distribution
qqPlot(residuals(Mcap_Color_sq))#view qq
plot(fitted(Mcap_Color_sq), residuals(Mcap_Color_sq))#view variance

#The data transformation chosen was (bleaching score +500)^2 
```

Mcapitata Model
```{r}
Mcap_Color_sq <- lmer(((Bleaching.Score+500)^2) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=mcap.color) #set model

## significance of model
Anova(Mcap_Color_sq, ddf="lme4", type='III') # Type III Wald chisquare tests
```

Testing Transformations for P acuta
```{r}
#Our random factor is `(1|PLUG.ID)` because fragments were repeatedly measured
Pacuta_Color_sqrt <- lmer(((Bleaching.Score+500)^(1/2)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=pacuta.color) 
par(mfrow = c(2, 2))
hist(residuals(Pacuta_Color_sqrt))#view distribution
boxplot(residuals(Pacuta_Color_sqrt))#view distribution
qqPlot(residuals(Pacuta_Color_sqrt))#view qq
plot(fitted(Pacuta_Color_sqrt), residuals(Pacuta_Color_sqrt))#view variance

Pacuta_Color_qrt <- lmer(((Bleaching.Score+500)^(1/4)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=pacuta.color)
par(mfrow = c(2, 2))
hist(residuals(Pacuta_Color_qrt))#view distribution
boxplot(residuals(Pacuta_Color_qrt))#view distribution
qqPlot(residuals(Pacuta_Color_qrt))#view qq
plot(fitted(Pacuta_Color_qrt), residuals(Pacuta_Color_qrt))#view variance

Pacuta_Color_log <- lmer(log10((Bleaching.Score+500)) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=pacuta.color)
par(mfrow = c(2, 2))
hist(residuals(Pacuta_Color_log))#view distribution
boxplot(residuals(Pacuta_Color_log))#view distribution
qqPlot(residuals(Pacuta_Color_log))#view qq
plot(fitted(Pacuta_Color_log), residuals(Pacuta_Color_log))#view variance

Pacuta_Color_sq <- lmer(((Bleaching.Score+500)^2) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=pacuta.color)
par(mfrow = c(2, 2))
hist(residuals(Pacuta_Color_sq))#view distribution
boxplot(residuals(Pacuta_Color_sq))#view distribution
qqPlot(residuals(Pacuta_Color_sq))#view qq
plot(fitted(Pacuta_Color_sq), residuals(Pacuta_Color_sq))#view variance

#The data transformation chosen was (bleaching score +500)^2 
```

Pacuta Model
```{r}
Pacuta_Color_sq <- lmer(((Bleaching.Score+100)^2) ~ Temperature*CO2*Timepoint + (1|PLUG.ID), na.action=na.omit, data=pacuta.color) #set model

## significance of model
Anova(Pacuta_Color_sq, ddf="lme4", type='III') # Type III Wald chisquare tests
```


Plotting color score for both species together
```{r}
Blch_plot <- ggplot(Blch.meanse, aes(x=Timepoint, y = mean, group = Treatment, color = Treatment)) +
  geom_line() + geom_point() + #plot points and lines
  facet_grid(~Species, scales = "free", labeller = species_labeller) + #facet by Species
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.2) +  #plot sem error bars
  scale_color_manual(values = cols) + #color points
  xlab("Timepoint") + #label x axis
  ylab("Color Score") + # label y axis
  theme_classic() + #set theme
  theme(legend.position = c(0.1,0.3)) + #set legend position
  theme(legend.text = element_text(size = 7)) + #set legend text size
  theme(legend.title = element_text(size = 9)) + #set legend text title size
  geom_vline(xintercept = c(7.8), colour="black", linetype="dotted") + #add vertical line
  theme(plot.title = element_text(face = 'bold.italic', size = 12, hjust = 0)) + 
  scale_x_discrete(breaks = c("Week1", "Week2", "Week4", "Week6", "Week8", "Week12", "Week16"), labels = c("1 week", "2 week", "4 week", "6 week", "8 week", "12 week", "16 week")) + #set breaks
  theme(axis.text.x = element_text(angle = 60, vjust = 1.2, hjust = 1.3)) + #Set the text angle
  theme(panel.border = element_blank(),#set theme
  panel.grid.major = element_blank(),#set theme
  panel.grid.minor = element_blank(),#set theme
  # Change axis line
  axis.line = element_line(colour = "black"))+ 
  theme(strip.text = element_text(face = "bold.italic")) 
Blch_plot #view plot

ggsave(file="Output/Color_Score.pdf", Blch_plot, width = 11, height = 6, units = c("in"))
ggsave(file="Output/Color_Score.png", Blch_plot, width = 11, height = 6, units = c("in"))


Mcap.info <- as.character("Montipora capitata")
Mcap.Stats <- Anova(Mcap_Color_sq, ddf="lme4", type='III') # Type III Wald chisquare tests
Pacuta.info <- as.character("Pocillopora acuta")
Pacuta.Stats <- Anova(Pacuta_Color_sq, ddf="lme4", type='III') # Type III Wald chisquare tests

#Capture statistical results to file
capture.output(Mcap.info, Mcap.Stats,  Pacuta.info, Pacuta.Stats, file="Output/Color_Score_Stat_Results.txt")


```


